Analysing Data from a Personal Activity Monitor Device 
=======================================================

We will first being by uncompressing and loading the data from `activity.zip` which is included with this document:

```{r message=FALSE}
library(dplyr)
library(lubridate)
unzip("activity.zip")
activity <- read.csv("activity.csv")
activity$date <- ymd(activity$date)
```

A quick summary of the data will yield the following:

```{r}
summary(activity)
```

```{r echo=FALSE}
num_nas <- sum(is.na(activity$steps))
```

Since we have found a large number of observations with missing step values (`r num_nas` with NAs), we will filter those out :

```{r}
activity_complete <- filter(activity, !is.na(steps))
```


##What is mean total number of steps taken per day?

To answer this question, we will first plot the histogram of the total number of steps taken each day:

```{r message=FALSE}
library(lattice)
day_grouped <- group_by(activity_complete, date)
total_steps_day <- summarise(day_grouped, total_per_day = sum(steps))
histogram(~total_steps_day$total_per_day, xlab="Steps per day")
mean(total_steps_day$total_per_day)
median(total_steps_day$total_per_day) 
```


##What is the average daily activity pattern?

We first begin by charting the average steps taken by this person for every interval in the day, for all the days sampled: 

```{r}
interval_grouped <- group_by(activity_complete, interval)
avg_steps_day <- summarise(interval_grouped, avg_day = sum(steps)/n())
xyplot(avg_day ~ interval, data=avg_steps_day, type="l", xlab="Time interval", ylab="Average Steps")
```

To see when this person is mostly active, we can look at the time interval where this person usually logs the most number of steps during the day:

``` {r}
filter(avg_steps_day, avg_day == max(avg_day))
```


## Imputing missing values

As mentioned in the initial section, the total number of NAs found in the raw dataset was `r num_nas`. 

To complete all the observations, We will replace missing values with the average of the time interval:

``` {r}
activity_estimated_nas <- activity
for (i in 1:nrow(activity_estimated_nas))
{
  row <- activity_estimated_nas[i, ]
  if (is.na(row["steps"]))
      {
        matching_interval <- filter(avg_steps_day, interval == row[["interval"]])
        row["steps"] <- matching_interval$avg_day
        activity_estimated_nas[i, ] <- row
      }
}
```


```{r}
sum(is.na(activity_estimated_nas$steps))
```

After filling the NAs, if we re-run the analysis to get the mean and average of the total steps taken we get:

```{r message=FALSE}
day_grouped_adj <- group_by(activity_estimated_nas, date)
total_steps_day_adj <- summarise(day_grouped_adj, total_per_day = sum(steps))
histogram(~total_steps_day_adj$total_per_day, xlab="Steps per day")
mean(total_steps_day_adj$total_per_day)
median(total_steps_day_adj$total_per_day) 
```

The difference between the dataset wihtout the NA values and the dataset with the imputted values:

```{r}
mean(total_steps_day$total_per_day) - mean(total_steps_day_adj$total_per_day)
median(total_steps_day$total_per_day)  - median(total_steps_day_adj$total_per_day) 
```


##Are there differences in activity patterns between weekdays and weekends?

```{r}
activity_day <- mutate(activity_estimated_nas, day_type=ifelse (weekdays(activity_estimated_nas$date) == "Saturday" | weekdays(activity_estimated_nas$date) == "Sunday", "weekend", "weekday")) 

interval_grouped <- group_by(activity_day, interval, day_type)
avg_steps_day <- summarise(interval_grouped, avg_day = sum(steps)/n())
xyplot(avg_day ~ interval | day_type, data=avg_steps_day, type="l", xlab="Time interval", ylab="Average Steps", layout= c(1,2))

```